\documentclass{ximera}

\input{../../preamble.tex}

\author{Carl Stitz \and Jeff Zeager \and  Bobby Ramsey}
\license{CC-By-SA-NC}
\acknowledgement{http://www.stitz-zeager.com/}

\begin{document}

\begin{exercise}
	Solve the rational equation:
	\[ \frac{x^3-2x+1}{x-1} = \frac{1}{2} x - 1 \]
	
	\[ \text{How many solutions are there? } \quad \answer{2} \]
	\begin{exercise}
		List them in order from left to right.
		\[ x = \answer{-1/2}  \quad \text{ and } \quad x=\answer{0} \]
	\end{exercise}
\end{exercise}
\begin{hint}
	Try rewriting the right-hand side as a fraction and cross-multiplying to clear fractions.
\end{hint}

\begin{exercise}
	Solve the rational equation:
	\[ \frac{1}{x+3} + \frac{1}{x-3} = \frac{x^2-3}{x^2-9} \]
	
	\[ \text{How many solutions are there? } \quad \answer{1} \]
	\begin{exercise}
		It is $x = \answer{-1}$.
		\begin{exercise}
			Why is $x=3$ NOT a solution?
			\begin{multipleChoice}
				\choice{It makes both sides equal.}
				\choice{It does not make both sides equal.}
				\choice[correct]{It is not in the domain.}
			\end{multipleChoice}
		\end{exercise}
	\end{exercise}
\end{exercise}
\begin{hint}
	Make sure to check your possible solutions are actually solutions.
\end{hint}

\begin{exercise}
	Solve the rational equation:
	\[ \frac{x^2-2x+1}{x^3+x^2-2x} = 1 \]
	\[ \text{There are } \, \answer{0} \, \text{solutions.} \]
	\begin{feedback}
		What are you left with if you multiply across to clear the fractions, then combine like terms?  Are there any values of $x$ making that statement true?	
	\end{feedback}		
\end{exercise}

\begin{exercise}
	Solve the rational inequality.  Express your answer in interval notation.  (Write your intervals in order from left to right)
	\[ \frac{-x^3+4x}{x^2-9} \geq 4x  \]
	
	\[ \left( \answer{-\infty} \, , \, \answer{-3} \right) \cup \left[ \answer{-2\sqrt{2}} \, , \, \answer{0} \right] \cup \left[ \answer{2\sqrt{2}} , \answer{3} \right) \]
		
\end{exercise}


\end{document}
