\documentclass{ximera}

\input{../preamble.tex}

\outcome{Review derivatives.}

\title[Dig-In:]{Review Derivatives}


\begin{document}
\begin{abstract}
  Review differentiation.
\end{abstract}
\maketitle


\section{What are derivatives?}
After working with limits, our main tool last semester was the derivative.  We started with a function $f$ and a number $a$ in the
interior of its domain.  The derivative of $f$ at $a$ ( denoted as either $f'(a)$ or $\eval{ \ddx f(x)}_{x=a}$ ) can be interpreted
in a couple different ways.  It represents the \emph{instantaneous rate of change of $f$ at $a$}.  Graphically it represents
the \emph{slope of the line tangent to the graph of $y=f(x)$ at the point $(a, f(a) )$}. 

When we first came up with the idea, we didn't know how to calculate $f'(a)$ explicitly, so we approximated instead.  We
approximated the \emph{instantaneous rate of change at $a$} by the \emph{average rate of change on $[a, a+h]$}.  Graphically
we approximated the \emph{slope of the tangent line at $(a,f(a))$} by the \emph{slope of the secant line through $(a, f(a))$ and $(a+h, f(a+h))$}.
The exact value arose when we took the limit as $h$ tended to $0$.

The definition of the derivative is then:
\begin{definition}
  The \dfn{derivative} of $f$ is 
  \[
  \ddx f(x) = \lim_{h\to 0} \frac{f(x+h) - f(x)}{h}.
  \]
\end{definition}

Let's go through an example of using this definition.
\begin{example}
	Find an equation of the line tangent to the graph of  $f(x) = x^2 - 3x + 1$ at the point $(2, -1)$.
	\begin{explanation}
		To find an equation for a line we need a point on the line and the slope of the line.  Since we know the point
		$(2, -1)$ is on the line, our equation should look like:
		\[ y + 1 = m_{tan} (x-2) \]
		where $m_{tan}$ is the slope.  
		
		We know the slope at $(2, -1)$ means we need to find $f'\left( \answer[given]{2} \right)$.
		\begin{align*}
			f'(2) &= \lim_{h \to 0} \frac{ f(2+h) - f(2) }{h}\\
				&= \lim_{h \to 0} \frac{\left((2+h)^2 - 3(2+h) + 1\right) - \left(  -1 \right)}{h}\\
				&= \lim_{h \to 0} \frac{\left( (4 + 4h + h^2)-(6+3h) + 1\right) +1}{h}\\
				&= \lim_{h \to 0} \frac{h^2+h}{h}\\
				&= \lim_{h \to 0} \frac{h(h+1)}{h}\\
				&= \lim_{h \to 0} \frac{\cancel{h}(h+1)}{\cancel{h}}\\
				&= \lim_{h \to 0} (h+1)\\
				&= \answer{1}
		\end{align*}

		An equation for the line is: \[ y + 1 = 1 (x-2). \]
	\end{explanation}
\end{example}

\section{Shortcut formulas}
We soon turned to finding `shortcut formulas' for finding derivatives more easily.   The most important of these are below.
\begin{description}
	\item[Power Rule] $\displaystyle \ddx\left( x^n \right) = n x^{n-1}$
	\item[Sum Rule] $\displaystyle \ddx \left( f(x) + g(x) \right) = f'(x) + g'(x)$
	\item[Product Rule] $\displaystyle \ddx \left( f(x) g(x) \right) = f'(x) g(x) + f(x) g'(x)$
	\item[Quotient Rule] $\displaystyle \ddx\left( \dfrac{f(x)}{g(x)} \right) = \dfrac{f'(x)g(x) - f(x) g'(x)}{(g(x))^2}$
	\item[Chain Rule] $\displaystyle \ddx \left( f\left(g(x) \right) \right) = f'\left( g(x) \right) g'(x)$
	\item[Exponential and Logarithmic Functions]{
		\begin{align*}
			\ddx \left( e^x \right) 		&= e^x\\
			\ddx \left( \ln x \right) 		&= \dfrac{1}{x}\\
			\\
			\ddx \left( b^x \right) 		&= b^x \ln b\\
			\ddx \left( \log_b x\right) 	&= \dfrac{1}{x \ln b}
		\end{align*}
	}	
	\item[Trigonometric Functions]{
		\begin{align*}
			\ddx\left( \sin x\right) 		&= \cos x\\
			\ddx \left( \cos x \right) 	&= - \sin x\\
			\ddx\left( \tan x \right) 	&= \sec^2 x\\
			\ddx \left( \sec x \right) 	&=  \sec x \tan x\\
			\ddx \left( \csc x \right) 	&= - \csc x \cot x\\
			\ddx \left( \cot x \right) 	&= - \csc^2 x
		\end{align*}	
	}		
\end{description}





\begin{example}
	If an object has displacement function given by $s(t) = 2t^2 - 4t + e^t$, find the velocity at time $t = 2$.
	\begin{explanation}
		Since velocity is the rate of change of displacement, this is asking us to find $s'(2)$.  We'll find
		$v(t) = s'(t)$ by differentiating term-by-term.
		\begin{align*}
			v(t) 	&= \ddt\left(2t^2-4t+e^t \right)\\
				&= \ddt\left(2t^2\right) - \ddt\left(4t\right) + \ddt \left( e^t \right)\\
				&= \answer{4t} - \answer{4} + \answer{e^t}
		\end{align*}
		This gives a formula for $v(t)$, but we are being asked for the velocity at a specific time.  We still need to
		plug in $t = 2$.
		\[ v(2) = 8 - 4 + e^2 = 4+ e^2. \]
	\end{explanation}
\end{example}

\begin{example}
	Find $\ddx \left( 2x \tan \left(\log_5 (x+2)\right) \right)$.
	\begin{explanation}
		At its outer-most level, this is a multiplication problem.  The $2x$ is being multiplied by a tangent function.  That means we'll start with
		the Product Rule.
		\begin{align*} 
		\ddx \left( 2x \tan\left( \log_5 (x+2) \right) \right) &= \ddx(2x) \tan\left( \log_5 (x+2) \right) \\
			&+ (2x) \ddx\left( \tan\left( \log_5 (x+2)\right) \right).
		\end{align*}
		The first term on we can handle, since $\displaystyle \ddx(2x) = \answer{2}$.  The term on the right asks us to differentiate a tangent function, but the stuff
		on the inside isn't just the variable itself.  That means it is a Chain Rule problem.
			\[ \ddx \left( \tan \left( \log_5 (x+2) \right) \right) = \sec^2\left( \log_5 (x+2) \right) \cdot \ddx \left( \log_5 (x+2) \right). \]
		We have to take the derivative of a logarithm, but $x$ is not alone on the inside.  There is ANOTHER Chain Rule.
			\[ \ddx \left( \log_5 (x+2) \right) = \dfrac{1}{(x+2) \ln( \answer[given]{5} )} \cdot \ddx(x+2) = \dfrac{1}{(x+2) \ln 5}. \]

		Putting this all together, our derivative is:
		\[ 2 \tan \left( \log_5 (x+2) \right) + \dfrac{2x \sec^2 \left( \log_5 (x+2) \right)}{(x+2) \ln 5}. \]
	\end{explanation}
\end{example}


\begin{example}
	Suppose $f$ is a function whose values are given in the following table.
	\begin{center}
		\begin{tabular}{c c c}
			\hline \hline
			x & f(x) & f'(x) \\ 
			\hline
			0 & 1 & -1 \\
			1 & -1 & 3\\
			2 & 6 & -7\\
			\hline
		\end{tabular}		
	\end{center}
	
	Find $\displaystyle \eval{ \ddx\left( \cos\left( (x+\pi) f( x ) \right) \right)}_{x=0}$.
	\begin{explanation}
		Notice that, at its outer-most level, this function is just the cosine of something.  That means this is a Chain Rule problem.  Let's start with that.
		\[\ddx\left( \cos\left( (x+\pi) f( x ) \right) \right) = -\sin\left( (x+\pi) f( x ) \right) \cdot \ddx\left( (x+\pi) f( x ) \right). \]
		
		The stuff that was on the inside of the cosine is a product of $(x + \pi)$ with $f(x)$, so next is Product Rule.
		\begin{align*} 
			\ddx \left( x + \pi\right) f(x) &= \ddx\left(x+ \pi\right) f(x) + (x + \pi) \ddx \left( f(x) \right)\\
				&= \left(\answer[given]{1}\right) f(x) + (x+\pi) f'(x).
		\end{align*}
		Altogether we have:
		\[\ddx\left( \cos\left( (x+\pi) f( x ) \right) \right) = -\sin\left( (x+\pi) f( x ) \right) \cdot \left(f(x) + (x+\pi) f'(x) \right). \]
		
		Plugging in $x=0$ gives:
		\begin{align*}
			\eval{\ddx\left( \cos\left( (x+\pi) f( x ) \right) \right)}_{x=0} &= -\sin\left( \pi f( 0 ) \right) \cdot \left(f(0) + \pi f'(0) \right) \\
				&= -\sin\left( \pi \right) \cdot \left( 1 - \pi \right)\\
				&= \answer{0}.
		\end{align*}
	\end{explanation}
\end{example}


\section{Applications}
We finished last semester with a few applications that showed some of the things the derivative was good for.  (There are MANY others!)

Recall that a \emph{critical point} of a function $f$ is a number $c$  with either $f'(c)=0$ or $f'(c)$ does not exist.
\begin{question}
	Find the critical points of the function $\displaystyle f(x) = x e^x$.
	
	\begin{prompt} \[x = \answer{-1}\]\end{prompt}
\end{question}

Critical points played a large role for us in graphing functions and in optimization problems. Try the following optimization problem.
\begin{question}
	Among all rectangles with perimeter $30$, find the maximum area possible.
	
	
	\begin{prompt} \[ \textrm{The maximum area is } \answer{\frac{225}{4}}. \] \end{prompt}
\end{question}

\end{document}
