\documentclass{ximera}
\input{../../preamble.tex}
\outcome{Calculate the limit as $x$ approaches $\pm \infty$ of common functions algebraically.}
\outcome{Compute limits at infinity of famous functions.}
\outcome{Compute slant asymptotes.}
\tag{limit}
\author{Bobby Ramsey}
\license{Creative Commons 3.0 By-NC}
\begin{document}

\begin{exercise}
	Let \[ g(x) = \frac{3x^4+x^3+2x+5}{x^3+2}. \]
	
	Compute

	\begin{align*}
		\lim_{x\to \infty} g(x) &= \answer{\infty}\\ \\
		\lim_{x\to -\infty}g(x) &= \answer{-\infty}
	\end{align*}

	\begin{hint}
		Multiply by \[\frac{\frac{1}{x^3}}{\frac{1}{x^3}}\]
	\end{hint}

	\begin{exercise}
		Does the function $g$ have any horizontal asymptotes?
		\begin{multipleChoice}
			\choice[correct]{No.}
			\choice{Yes.}
			\choice{We cannot tell with what we are given.}
		\end{multipleChoice}
		\begin{hint}
			What do you get if you divide $3x^4+x^3+2x+5$ by $x^3+2$?
		\end{hint}
		
		\begin{exercise}
			What is the quotient and remainder if $3x^4+x^3+2x+5$ is divided by $x^3+2$?
			\begin{align*}
				\text{ Quotient } &= \answer{3x+1}\\
				\text{ Remainder } &= \answer{-4x+3}	
			\end{align*}
			\begin{exercise}
				Write $g(x)$ as a proper fraction:
				\[ g(x) = \left( \answer{3x+1} \right) + \frac{\answer{-4x+3}}{x^3+2} \]
				\begin{exercise}
					What is the end behavior of the fraction $\dfrac{-4x+3}{x^3+2}$?
					\begin{align*}
						\lim_{x\to \infty} \frac{-4x+3}{x^3+2} &= \answer{0} \\ \\
						\lim_{x\to -\infty} \frac{-4x+3}{x^3+2} &= \answer{0} \\
					\end{align*}
					\begin{exercise}
						The function $g$ did not have any horizontal asymptotes.  That means it did not approach any horizontal lines as $x$ went off toward either $\pm \infty$.
						There is another line that $g$ approached, though.  This is the \wordChoice{\choice[correct]{slant asymptote}\choice{tangent line}\choice{secant line}}.
						\begin{exercise}	
							The equation of the slant asymptote is:
							\[ y = \answer{3x+1} \]
						\end{exercise}
					\end{exercise}
				\end{exercise}
			\end{exercise}
		\end{exercise}
	\end{exercise}
\end{exercise}
\end{document}
