\documentclass{ximera}

\input{../preamble.tex}

\outcome{Recognize a composition of functions.}
\outcome{Take derivatives of compositions of functions using the chain rule.}
\outcome{Take derivatives that require the use of multiple derivative rules.}
\outcome{Use the chain rule to calculate derivatives from a table of values.}
\outcome{Understand rate of change when quantities are dependent upon each other.}
\outcome{Use order of operations in situations requiring multiple derivative rules.}
\outcome{Apply chain rule to relate quantities expressed with different units.}

\title[Break-Ground:]{An unnoticed composition}

\begin{document}
\begin{abstract}
Two young mathematicians discuss the chain rule.
\end{abstract}
\maketitle

Check out this dialogue between two calculus students (based on a true
story):

\begin{dialogue}
\item[Devyn] Riley! Something is bothering me. 
\item[Riley] What is it?
\item[Devyn]  It's about rates of change.
\item[Riley] That's just a derivative.  What's the issue?
\item[Devyn]  It might take a bit of setup.  Suppose my car has a 10 gallon tank.  The distance $D$ I can drive, without filling up, is $D(m)=10m$ miles, where $m$ is my car's fuel
			efficiency, in miles per gallon.
\item[Riley]  Sure.  So if you're getting $35$ miles per gallon, that is $D(35) = 350$ miles, so you can drive 350 miles before running out of gas.
\item[Devyn] Right!  Now, the fuel efficiency depends on how fast I'm driving.  If I'm driving 55 miles per hour, I can get 40 miles per gallon, but if I'm driving 70 miles per hour on the interstate,
			I only get 30 miles per gallon.
\item[Riley]  Oh!  So your distance function $D$ depends on $m$, but $m$ depends on your velocity $v$.  That means $D$ is really a function of $v$.
\item[Devyn]  Exactly!  Finding the derivative $\frac{dD}{dm}$ is easy, but how do I find $\frac{dD}{dv}$?
\item[Riley]  If only we could find the derivative of a composite function...
\end{dialogue}



\begin{question}
	What is $\frac{dD}{dm}$?
	\[ \frac{dD}{dm} = \answer{10}. \]
\end{question}

\begin{question}
	Suppose that the fuel efficiency $m$ is a linear between $v = 50$ mph to $v= 70$ mph.  What is $\frac{dm}{dv}$?
	\begin{multipleChoice}
		\choice{$10$}
		\choice{$-10$}
		\choice[correct]{$-1.5$}
		\choice{$1.5$}
	\end{multipleChoice}
\end{question}

%\input{../leveledQuestions.tex}


\end{document}
