\documentclass{ximera}

\input{../preamble.tex}

\outcome{Know the graphs and properties of ``famous'' functions.}

\title[Dig-In:]{Working with polynomials}


\begin{document}
\begin{abstract}
  Polynomials are some of our favorite functions. 
\end{abstract}
\maketitle


The functions you are most familiar with are probably polynomial
functions.

\section{What are polynomial functions?}

\begin{definition}
  A \dfn{polynomial function} in the variable $x$ is a function
  which can be written in the form
  \[
  f(x) = a_nx^n + a_{n-1}x^{n-1} + \dots + a_1 x + a_0
  \]
  where the $a_i$'s are all constants (called the \dfn{coefficients})
  and $n$ is a whole number (called the \dfn{degree} when $n\ne
  0$). The domain of a polynomial function is $(-\infty,\infty)$.
\end{definition}

\begin{question}
	Which of the following are polynomial functions?
	\begin{selectAll}
		\choice[correct]{$f(x) = 0$}
		\choice[correct]{$f(x) = -9$}
		\choice[correct]{$f(x) = 3x+1$}
		\choice{$f(x) = x^{1/2}-x +8$}
		\choice{$f(x) = -4x^{-3}+5x^{-1}+7-18x^2$}
		\choice[correct]{$f(x) = (x+1)(x-1)+e^x - e^x $}
		\choice{$f(x) = \frac{x^2 - 3x + 2}{x-2}$}
		\choice[correct]{$f(x) = x^7-32x^6-\pi x^3+45/84$}
	\end{selectAll}
\end{question}

The phrase above ``in the variable $x$'' can actually change.
\[
y^2-4y +1
\]
is a polynomial in $y$, and
\[
\sin^2(x) + \sin(x) -3 
\]
is a polynomial in $\sin(x)$.


\section{Multiplying}

Multiplying polynomials is based on the familiar property of arithmetic, \dfn{distribution}: $\displaystyle a\cdot( b + c ) = ab + ac$.
\begin{example}
	Multiply $2x^3$ by $x^5 - 4x^4 + 7x - 2$.
	\begin{explanation}
		Use distribution.
		\begin{align*}
			2x^3 \left( x^5 - 4x^4 + 7x - 2 \right) &= 2x^3 \cdot x^5 - 2x^3 \cdot 4x^4 + 2x^3 \cdot 7x \\
				& - 2x^3 \cdot 2\\
				&= 2 x^8 - 8x^7 + 14x^4 - 4x^3
		\end{align*}
	\end{explanation}
\end{example}


\begin{example}
	Multiply $3x^2 + 2x - 1$ by $2x^4 + 5x^3+ x + 1$.
	\begin{explanation}
		We'll start by distributing the first polynomial into the second.  Then we'll distribute back, and finish by combining like terms.
		\begin{align*}
			(3x^2+ 2x-1)(2x^4 + 5x^3 + x +1) &= 2x^4(3x^2+ 2x-1) \\
			 &+ 5x^3 (3x^2+ 2x-1) \\
			 &+ x (3x^2+ 2x-1) \\
			 &+ 1(3x^2+ 2x-1)\\
				&= (6x^6 + 4x^5 - 2x^4) \\
				&+ (15x^5 + 10x^4 - 5x^3) \\
				&+ (3x^3+2x^2-x) \\
				&+(3x^2+2x-1)
		\end{align*}
		After combining like-terms, we find \[6x^6 + 19x^5 +8x^4 - 2x^3 + 5x^2+ x - 1.\]
	\end{explanation}
\end{example}
The result is that we have multiplied every term of the first polynomial by every term of the second, then added the results together.  In the case of two
\dfn{binomials} (polynomials with only two terms), this is frequently referred to as \dfn{FOIL}.

There are several product formulas that arise repeatedly when working with binomials.  You will likely have seen most of these before.
%\begin{xbox}{Multiplication Formulas}
		\begin{align*}
			(a + b)^2 &= a^2 + 2ab + b^2\\
			(a-b)^2 &= a^2 - 2ab + b^2\\
			(a+b)(a-b) &= a^2 - b^2\\
			(a+b)^3 &= a^3 + 3 a^2 b + 3 ab^2 + b^3\\
			(a-b)^3 &= a^3 - 3 a^2 b + 3 ab^2 - b^3\\
			(a-b)(a^2 + ab + b^2) &= a^3 - b^3\\
			(a+b)(a^2-ab+b^2) &= a^3 + b^3
		\end{align*}
%\end{xbox}


\begin{example}
	If $f$ is the polynomial function given by $f(x) = x^3-2x^2 +5$, find $f(x+1)$.  Find $f(2+h)$.
	\begin{explanation}
		$f(x+1)$ means to replace $x$ in the formula for $f(x)$ by $x+1$.
		\begin{align*}
			f(x+1) &= (x+1)^3 - 2(x+1)^2 + 5 \\
				&= (x^3 + 3x^2 + 3x + 1) - 2(x^2 + 2x + 1) + 5\\
				&= x^3 + x^2 -x + 4
		\end{align*}
		In the same way, $f(2+h)$ asks us to replace $x$ by $2+h$.
		\begin{align*}
			f(2+h) &= (2+h)^3 - 2(2+h)^2 + 5\\
				&= (8+12h + 6h^2 + h^3) - 2(4+4h+h^2) + 5\\
				&= h^3 + 4h^2+4h+5
		\end{align*}
	\end{explanation}
\end{example}

\begin{problem}
	If $f(x)=4x^2-3x+1$, find $f(x+h)-f(x)$.
	\begin{multipleChoice}
		\choice{$h$}
		\choice{$4h^2-3h+1$}
		\choice[correct]{$8xh + 4h^2 - 3h$}
		\choice{$4x^2+8xh+4h^2-3x-3h+1$}
	\end{multipleChoice}
\end{problem}

\section{Factoring}
Factoring is a bit like an inverse operation of multiplying polynomials.  We start with the multiplied out polynomial, and ask what the individual factors were.

The easiest factors to deal with are common factors.
\begin{example}
	Factor $8x^5 + 40x^4 + 40x^3$.
	\begin{explanation}
		The coefficients of this polynomial have a greatest common divisor of $8$, and the least degree of the terms is $3$, so
		$8x^3$ is a common factor.  Taking this out leaves:
		$\displaystyle 8x^3( x^2+5x+5)$.
	\end{explanation}
\end{example}


For \dfn{trinomials} (polynomials with three terms) of the form $ax^2 + bx + c$, we try to factor as $(px+r)(qx+s)$.  We'll start with an example with $a=1$.
\begin{example}
	Factor $x^2+2x-24$.
	\begin{explanation}
		Since the leading coefficient is $1$, we want to factor $x^2+2x-24$ as $(x+r)(x+s)$.  If we multiply out $(x+r)(x+s)$ we get
		$x^2 + (r+s)x + rs$.  We need to find two numbers that add to $2$,
		and multiply to $-24$.  Look at the different ways of multiplying two numbers to get $-24$:  $-1 \cdot 24$, $-2 \cdot 12$, $-3 \cdot 8$, $-4 \cdot 6$...
		That last one, $-4$ and $6$ are two numbers that add to $2$ and multiply to give $-24$.
		
		Our factorization is $x^2+2x-24 = (x-4)(x+6)$.
	\end{explanation}
\end{example}

The process is slightly more complicated when $a \neq 1$.
\begin{example}
	Factor $6x^2 + 11 x + 3$.
	\begin{explanation}
		In this case, want to find $(px+r)(qx+s) = pq x^2 + (ps+qr) x+ rs$.  Ignore that middle term for now, and just focus on the leading term and the constant term.
		We need to find numbers that multiply to $6$ for our coefficients, and that multiply to $3$ for the constants in each factor.
		That means we're looking at things like $(6x+1)(x+3)$, $(6x+3)(x+1)$, $(2x+3)(3x+1)$, $(2x+1)(3x+3)$.   Actually, that's it, as $6$ can only
		factor as $1 \cdot 6$ or $2 \cdot 3$, and $3$ only factors as $1 \cdot 3$.
		
		All of these possibilities will give the right $x^2$ term and the right constant term.  The only difference is the $x$-term they give.  Do any of them
		give an $x$-coefficient of $11$?  Yes!
		
		$6x^2+11x+3 = (2x+3)(3x+1)$.
	\end{explanation}
\end{example}

Remember, a \dfn{root} of a polynomial function is an $x$-value where the polynomial is zero. 
There is a close relationship between roots of a polynomial and factors of that polynomial.  Precisely:
\begin{theorem}
If $x=c$ is a root of a polynomial, then $x-c$ is a factor of that polynomial.
\end{theorem}

\begin{example}
	Factor $3x^3-13x^2+2x+8$.
	\begin{explanation}
		Notice that if we substitute $x=1$ into the polynomial, we get zero out.  That means $x-1$ is a factor of $3x^3-12x^2+2x+8$.
		Polynomial long division gives:
		\[\polylongdiv{3x^3-13x^2+2x+8}{x-1}\]
		The quotient was $3x^2-10x-8$, so $3x^3-13x^2+2x+8 = (x-1)(3x^2-10x-8)$.  It remains to factor the quotient.
		
		The leading coefficient $3$ factors as $1 \cdot 3$, and $8$ factors as $1 \cdot 8$ and $2 \cdot 4$.  Examining the possibilities, we find the factorization
		of $3x^2-10x-8$ as $(3x+2)(x-4)$.  The entire polynomial factors as $(x-1)(3x+2)(x-4)$.
	\end{explanation}
\end{example}



\section{Equations}
A \emph{quadratic equation in $x$} is an equation which is equivalent to one with the form
	$ax^2 + bx + c = 0$, where $a$, $b$, and $c$ are constants, with $a \neq 0$.

There are three major techniques you are probably familiar with for solving quadratic equations:
\begin{itemize}
	\item Factoring.
	\item Completing the Square.
	\item Quadratic Formula.
\end{itemize}

Each of these methods are important.  Factoring is vital, because it is a valid approach to solve nearly any type of equation.  Completing
the Square is a technique that becomes useful when we need to rewrite certain types of expressions.  The Quadratic Formula will always
work, but has some limitations.

\begin{example}
	Solve the quadratic equation
	\[ 2x^2 +5x = 27 x - 60. \]
	\begin{explanation}
		We'll start by writing this in it's standard form: $2x^2 - 22x + 60 = 0$.  Notice how there is a common factor of 2? Let's divide by 2.
		
		$x^2 - 11x + 30 = 0$.  Any of the above methods will work here, so let's try factoring.  What numbers add to $-11$ and multiply to $30$?
		$-5$ and $-6$ do. That gives us:
		\begin{align*}
			x^2 - 11x + 30 &= 0\\
			(x-5)(x-6) &= 0
		\end{align*}
		Either $x-5 = 0$ (giving us $x=5$) or $x-6 = 0$ (giving us $x=6$).  The two solutions are $x = 5, 6$.
	\end{explanation}
\end{example}	

\begin{example}
	Solve the quadratic equation
	\[  x^2 = 6x -4.\]
	\begin{explanation}
		Again, we'll write it in standard form: $x^2 - 6x + 4 = 0$.  The quadratic $x^2 -6x+4$ does not factor, so we'll complete the square instead.
		
		Start by moving the constant term to the other side
		\[ x^2 - 6x = -4. \]
		This has $b = -6$.  That means $\displaystyle \dfrac{b}{2} = -3$ and $\displaystyle \left( \dfrac{b}{2}\right)^2 = 9$.  Add $9$ to both sides.
		\[ x^2 - 6x + 9 = -4 + 9 = 5. \]
		The left-hand side is now a perfect square, $\left(x-3\right)^2 = x^2 - 6x + 9$.  We can then solve using square roots.
		\begin{align*}
			x^2 - 6x + 9 &=  5\\
			\left( x-3 \right)^2 &= 5\\
			x-3 = \pm \sqrt{5} \\
			x = 3 \pm \sqrt{5}.
		\end{align*}
	\end{explanation}
\end{example}
If you had used the quadratic formula, $\displaystyle x = \dfrac{-b \pm \sqrt{b^2-4ac}}{2a}$, instead of factoring or completing the square above, you would have found the same solutions.

\begin{problem}
	Solve the quadratic equation $\displaystyle x^2 + 4 = 4\left(x+2\right)$.
	\begin{multipleChoice}
		\choice{$x = \pm 2$}
		\choice[correct]{$x=2\pm 2\sqrt{2}$}
		\choice{$x = -2, -4$}
		\choice{none of the above}
  \end{multipleChoice}
\end{problem}




Building on these methods, we'll consider more general types of equations.
A polynomial equation is an equation which is equivalent to one of the form $a_n x^n + a_{n-1} x^{n-1} + \dots + a_1 x + a_0 = 0$. 
\begin{example}
	Solve the equation $x^4 = 6x^2 - 4$.
	\begin{explanation}
		Start by rewriting as $x^4 - 6x^2 + 4=0$.  This is not a quadratic equation in $x$, but it IS a quadratic in $x^2$.
		We continue by completing the square.
		\begin{align*}
			x^4 - 6x^2 &= 4 \\
			x^4 - 6x^2 + 9 &= 4 + 9\\
			(x^2 - 3)^2 &= 13\\
			x^2 - 3 &= \pm \sqrt{13}\\
			x^2 &= 3 \pm \sqrt{13}
		\end{align*}
		Either $x^2 = 3 + \sqrt{13}$, or $x^2 = 3 - \sqrt{13}$.  Since $3 - \sqrt{13}$ is negative, only $x^2 = 3 + \sqrt{13}$ is possible.
		
		The only real solutions are $x = \pm \sqrt{3 + \sqrt{13}}$.
	\end{explanation}
\end{example}

Factoring is a process that helps us solve more than just quadratic equations, as long as we first get one side of the equation equal to zero.
\begin{example}
	Solve the equation
	\[ 2x^2 \left( x-4 \right) \left( 2x-5\right)^3 = 0. \]
	\begin{explanation}
		The only way for a product of real numbers to be zero, is for one of the factors to have been zero.
		Here, that means either $2x^2 = 0$, or $x-4 = 0$ or $(2x-5)^3 = 0$.
		This gives us three equations, each substantially less complicated than the original one, to solve.  By solving them, we find
		$x = 0, 4, \frac{5}{2}$.
	\end{explanation}
\end{example}

\begin{example}
	Solve  $2x(2x^2 + 1) = x^2(x +13) - 8$.
	\begin{explanation}
		Start by rewriting the equation as $3x^3-13x^2+2x+8 = 0$.  This is not a quadratic-type equation, so we resort to factoring.
		\begin{align*}
			3x^3-13x^2+2x+8 &= 0\\
			(x-1)(3x^2-10x-8) &= 0\\
			(x-1)(3x+2)(x-4) &= 0
		\end{align*}
		Thus, either $x -1 = 0$ (so $x = 1$), or $3x+2=0$ (so $x = -2/3$), or $x-4 = 0$ (so $x=4$).
		
		The solutions are $x = 1, 4, -3/2$.
	\end{explanation}
\end{example}


The next theorem above is a deep fact of mathematics. The great mathematician Gauss proved the theorem in 1799. 
\begin{theorem}[The Fundamental Theorem of Algebra]\index{Fundamental Theorem of Algebra}
  Every polynomial of the form
  \[
  a_n x^n + a_{n-1} x^{n-1} + \dots + a_1 x + a_0
  \]
  where the $a_i$'s are real (or even complex!) numbers and $a_n \ne 0$ has exactly
  $n$ (possibly repeated) complex roots.
\end{theorem}

Recall that the \dfn{multiplicity} of a root indicates how many times that particular root is repeated.
The Fundamental Theorem of Algebra tells us that a polynomial equation of degree $n$, will have exactly $n$ complex solutions, once multiplicity
is taken into account.  As non-real solutions appear in complex-conjugate pairs  (if our equation has real coefficients), we will always have an even number of non-real
solutions.  That means, taking multiplicity into account, a quadratic equation will have either $2$ or $0$ real solutions.  A cubic equation will have either $3$ or $1$ real
solution.


\begin{problem}

	The equation $x^3 = 3x-2$ has $x=1$ as one solution.  Find another solution.
	
	\begin{prompt}
		\[ x= \answer{-2} \]
	\end{prompt}
	\begin{feedback}
  		Notice that there were only two real numbers that solve this equation.  How is that possible, if we said a cubic equation has to have either
		$3$ or $1$ real solutions?
  	\end{feedback}
\end{problem}



\section{End behavior}
We will shortly turn our attention to graphs of polynomial functions, but we have one more topic to discuss \emph{End Behavior}.
Basically, we want to know what happens to our function as our input variable $x$ gets really, really large in either the positive or negative direction.
This is kind of like the limits we talked about before, except $x$ is not approaching a fixed value $a$, but just going off to either the right or the left.

We'll start with the most basic polynomials, the \emph{monomials}, $x^n$.  (Since \emph{mono-} means one, the monomials are the polynomials with a single term.)
Here are the graphs of $f(x)=x^n$ for $n = 2$, $4$, and $6$:
	\begin{center}
		\begin{tikzpicture}
        			\begin{axis}[
          			domain=-2:2,
          			xmin=-2, xmax=2,
          			ymin=-8, ymax=8,
          			width=4in,
          			axis lines =middle, xlabel=$x$, ylabel=$y$,
          			every axis y label/.style={at=(current axis.above origin),anchor=south},
          			every axis x label/.style={at=(current axis.right of origin),anchor=west},
          			]
	 	 		\addplot [very thick, penColor, smooth] {x^2};
				\addplot [very thick, penColor2, smooth] {x^4};
				\addplot [very thick, penColor4, smooth] {x^6};
        			\end{axis}
      		\end{tikzpicture}
	\end{center}

	
Here are the graphs of $f(x)=x^n$ for $n = 3$, $5$, and $7$:
	\begin{center}
		\begin{tikzpicture}
        			\begin{axis}[
          			domain=-2:2,
          			xmin=-2, xmax=2,
          			ymin=-8, ymax=8,
          			width=4in,
          			axis lines =middle, xlabel=$x$, ylabel=$y$,
          			every axis y label/.style={at=(current axis.above origin),anchor=south},
          			every axis x label/.style={at=(current axis.right of origin),anchor=west},
          			]
	 	 		\addplot [very thick, penColor, smooth] {x^3};
				\addplot [very thick, penColor2, smooth] {x^5};
				\addplot [very thick, penColor4, smooth] {x^7};
        			\end{axis}
      		\end{tikzpicture}
	\end{center}

Notice the similarity between the all the graphs with $n$ even.  They all have the same basic cup shape, but higher values of $n$ make the graphs flatter in $(-1,1)$ and steeper outside $(-1,1)$.

The same thing happens with the $n$ odd graphs.  They have the same basic shape, but higher values of $n$ make the graphs flatter near the origin and steeper past $1$ and $-1$.

When $n$ is even, what is happening to the output values of $f(x) = x^n$ as $x$ gets larger and larger?  They themselves get larger and larger!  As $x$ increases without bound, so do the outputs.
The same thing happens as $x$ gets larger and larger in the negative direction without bound.  This is the end behavior we were looking for.  We'll say it this way:

\begin{align*}
	\text{As} \quad x \to \infty \, ,  \quad & x^n \to \infty \\
	\text{As} \quad x \to -\infty \, , \quad & x^n \to \infty \quad \text{, for $n$ even}
\end{align*}

For $n$ odd we have a slightly different end behavior.
\begin{align*}
	\text{As} \quad x \to \infty \, ,  \quad & x^n \to \infty \\
	\text{As} \quad x \to -\infty \, , \quad & x^n \to -\infty \quad \text{, for $n$ odd}
\end{align*}

Next we need to see how a coefficient could change this.  Remember how multiplying by a constant transforms a graph?  If the constant is positive the graph is vertically stretched/compressed.
If the constant is negative the graph is flipped over the $x$-axis and then vertically stretched/compressed.  This means if the coefficient of $x^n$ is positive, the end behavior is unaffected.  If the 
coefficient is negative, the end behavior is negated as well.

\begin{example}
	Find the end behavior of $f(x) = -3x^4$.
	\begin{explanation}
		Since $4$ is even, the function $x^4$ has end behavior 
		\begin{align*}
			\text{As} \quad x \to \infty \, ,  \quad & x^4 \to \infty \\
			\text{As} \quad x \to -\infty \, , \quad & x^4 \to \infty 
		\end{align*}
		The coefficient is negative, changing our end behavior to
		\begin{align*}
			\text{As} \quad x \to \infty \, ,  \quad & -3x^4 \to -\infty \\
			\text{As} \quad x \to -\infty \, , \quad & -3x^4 \to -\infty 
		\end{align*}
	\end{explanation}
\end{example}

\begin{problem}
	Find the end behavior of $g(x) = -6 x^9$.
	\begin{align*}
		\text{As} \quad x \to \infty \, ,  \quad & -6x^9 \to \answer{-\infty} \\
		\text{As} \quad x \to -\infty \, , \quad & -6x^9 \to \answer{\infty} 
	\end{align*}
\end{problem}


We understand monomials.  What about more general polynomials? 
\begin{example}
	Find the end behavior of $f(x) = 3x^5 - 4x^3 + 1$.
	\begin{explanation}
		If we rewrite the function as:
		\begin{align*}
			f(x) &= 3x^5 - 4x^3 + 1 \\
				&= x^5 \left( \frac{3x^5}{x^5} - \frac{4x^3}{x^5} + \frac{1}{x^5} \right)\\
				&=x^5 \left( \frac{3x^5}{x^5} - \frac{4x^3}{x^5} + \frac{1}{x^5} \right)\\
				&= x^5 \left( 3 - \frac{4}{x^2} + \frac{1}{x^5} \right)
		\end{align*}
		Remember that as $x\to \pm\infty$ we saw that $x^2 \to \infty$ and that $x^5 \to \mp \infty$.  That means $\frac{4}{x^2} \to 0$ and $\frac{1}{x^5} \to 0$. 
		(We'll talk about this more when we talk about Rational Functions.)  This means everything in parentheses will basically be just $3$ as $x \to \pm \infty$.
		That is, the polynomial $3x^5 - 4x^3 + 1$ has the same end behavior as $3x^5$.  Thus
		\begin{align*}
			\text{As} \quad x \to \infty \, ,  \quad & f(x) \to \infty \\
			\text{As} \quad x \to -\infty \, , \quad & f(x) \to -\infty 
		\end{align*}
	\end{explanation}
\end{example}

This example showed us that \emph{the end behavior of a polynomial is the same as the end behavior of its leading term}.
\begin{problem}
	Find the end behavior of $g(x) = -6x^9 + 15x^5 + 7 x^4 - 18 x^3 + 91x^2 - 72 x + 4$
	\begin{align*}
		\text{As} \quad x \to \infty \, ,  \quad & g(x) \to \answer{-\infty} \\
		\text{As} \quad x \to -\infty \, , \quad & g(x) \to \answer{\infty} 
	\end{align*}
\end{problem}




\section{What do the graphs look like?}
Polynomials of degrees $1$ and $2$ are well understood.

Polynomial functions with degree $1$ are referred to as linear polynomials.  This is due to the fact
that such a function can be written as $f(x) = mx + b$.  The graph of such a function is a straight line
with slope $m$ and $y$-intercept at $(0,b)$.

Quadratic functions, written as $f(x) = ax^2 + bx + c$ with $a \ne 0$, have parabolas as their graphs.
\begin{example}
	Sketch the graph of the function $f(x) = \frac{1}{2} x^2 + 2x + 1$.
	\begin{explanation}
		To plot the parabola, we first complete the square to write our function in the form $f(x) = a(x-h)^2 +k$.
		\begin{align*}
			f(x) &= \frac{1}{2} x^2 + 2x + 1 \\
				&= \frac{1}{2} \left( x^2+ 4x \right) + 1\\
				&= \frac{1}{2}\left( x^2 + 4x + 4\right) + 1 - \frac{1}{2} \cdot 4\\
				&= \frac{1}{2} \left( x+2 \right)^2 -1
		\end{align*} 
		The parabola we are looking for has vertex at $(-2, -1)$, opens upward (since $1/2 > 0$), and is wider than the standard
		$y=x^2$ parabola.
		\begin{tikzpicture}
        			\begin{axis}[
          			domain=-6:2,
          			xmin=-6, xmax=2,
          			ymin=-2, ymax=8,
          			width=2.5in,
          			axis lines =middle, xlabel=$x$, ylabel=$y$,
          			every axis y label/.style={at=(current axis.above origin),anchor=south},
          			every axis x label/.style={at=(current axis.right of origin),anchor=west},
          			]
	 	 		\addplot [very thick, penColor, smooth] {0.5*x^2+2*x+1};
        			\end{axis}
      		\end{tikzpicture}
		
		In terms of graph transformations, we can think of this as the graph of $g(x) = x^2$, which has been shifted horizontally $2$ units to the left,
		vertically compressed by a factor of $\frac{1}{2}$, then shifted vertically $1$ unit downward.
	\end{explanation}
\end{example}


\begin{problem}
	Use the graph of $f(x) = 2x^2 - 4x + 3$ to estimate the value of $\lim_{x\to 2} f(x)$.
	\[ \begin{prompt} 
		\lim_{x\to 2} f(x) = \answer{3} 
	\end{prompt}\]
\end{problem} 

For polynomials of higher degree, we will have to take everything else we have been talking about and put it together.  
We need to understand the zeroes of the polynomial and its end behavior.  The zeroes will correspond to $x$-intercepts, and the
end behavior will tell us what happens out past those intercepts.

One upshot of the Fundamental Theorem of Algebra (in terms of graphing) is that when we plot a
polynomial of degree $n$, its graph will cross the $x$-axis \emph{at most}
$n$ times.  Each crossing corresponds to a real root of that polynomial.  (Complex roots do not give crossings!)

\begin{example}
	Sketch a rough graph of the function $f(x) = 2(x-1)^2(x+2)$.

	\begin{explanation}
		We start with the end behavior of this function.  It is a degree $3$ polynomial with leading coefficient $2$, so
		\begin{align*}
			\text{As} \quad x \to \infty \, ,  \quad & f(x) \to \infty \\
			\text{As} \quad x \to -\infty \, , \quad & f(x) \to -\infty 
		\end{align*}
		We also see from the factorization that $x=1$ is a zero of multiplicity $2$, and $x=-2$ is a zero of multiplicity $1$.
		That means our graph will have $x$-intercepts at those locations.
		
		At these $x$-intercepts, will the graph pass through the $x$-axis like a line does, or will it touch the axis and turn back
		around, like a parabola does at it's vertex?  To answer this, look at a sign-chart for $f$.

		\begin{center}
		\begin{tikzpicture} 
			\tkzTabInit[lgt=2,espcl=1] 
				{$x$         /1, 
				$2$   /1, 
				$\left(x-1\right)^2$  /1,
				$x+2$       /1}% 
				{  , $-2$ , $1$ ,  }% 
			\tkzTabLine{ , + , t , + , t , + , }
			\tkzTabLine{ , + , t , + , t , + ,}
			\tkzTabLine{ , - , t , - , t , + ,  }
		\end{tikzpicture} 
		\end{center}
		
		Notice that around $x=-2$, only the $x+2$ factor changes signs.  That means $f(x)$ will change signs, so the graph will pass through the $x$-axis at $x=-2$.
		At $x=1$, however, the $(x-1)^2$ factor does NOT change signs (due to the even exponent).  That means $f(x)$ will turn around at that $x$-intercept.  Putting
		this together, we have the following graph.
		
		\begin{center}
		\begin{tikzpicture}
        			\begin{axis}[
          			domain=-3:3,
          			xmin=-3, xmax=3,
          			ymin=-2, ymax=9,
          			width=3in,
          			axis lines =middle, xlabel=$x$, ylabel=$y$,
          			every axis y label/.style={at=(current axis.above origin),anchor=south},
          			every axis x label/.style={at=(current axis.right of origin),anchor=west},
          			]
	 	 		\addplot [very thick, penColor, smooth] {2*((x-1)^2)*(x+2)};
        			\end{axis}
      		\end{tikzpicture}
		\end{center}
	\end{explanation}
\end{example}

Our discussion in that example showed that, given an $x$-intercept, we can determine if the graph passes through the $x$-axis or just turns around there, by
examining the multiplicity of that zero.  
\begin{theorem}
	Suppose the polynomial function $f$ has a zero at $x=a$ of multiplicity $n$.  If $n$ is odd, then the graph of $f$ will pass through $x$-axis at the point $(a,0)$.
	If $n$ is odd, then the graph of $f$ will intersect the $x$-axis at the point $(a,0)$, but not pass through the $x$-axis.
	
	More specifically: The graph of $f$ will resemble, at least locally near the point $(a,0)$, of the graph of $\pm (x-a)^n$.
\end{theorem}
This last statement means that if $x=4$ is a zero of $f$ with multiplicity $3$, then the graph of $f$ will pass through the $x$-axis at $(4,0)$, but it will flatten out the same
way that the graph of $y=x^3$ flattens out around the origin.

\begin{example}
  Here we see the the graphs of four polynomial functions.
  \begin{image}
    \begin{tabular}{cc}
      \begin{tikzpicture}
        \begin{axis}[
          domain=-2:2,
          xmin=-2, xmax=2,
          ymin=-2, ymax=2,
          width=2.5in,
          axis lines =middle, xlabel=$x$, ylabel=$y$,
          every axis y label/.style={at=(current axis.above origin),anchor=south},
          every axis x label/.style={at=(current axis.right of origin),anchor=west},
          ]
	  \addplot [very thick, penColor, smooth] {5*x^5-5*x^4-5*x^3+5*x^2+.5*x -1};
          \node at (axis cs:1.2, 1 ) [penColor,anchor=west] {$A$};
        \end{axis}
      \end{tikzpicture}
      &
      \begin{tikzpicture}
        \begin{axis}[
          domain=-2:2,
          xmin=-2, xmax=2,
          ymin=-2, ymax=2,
          width=2.5in,
          axis lines =middle, xlabel=$x$, ylabel=$y$,
          every axis y label/.style={at=(current axis.above origin),anchor=south},
          every axis x label/.style={at=(current axis.right of origin),anchor=west},
          ]
	  \addplot [very thick, penColor2, smooth] {-5*x^5+5*x^4+5*x^3-4.25*x^2-.3*x +.5};
          \node at (axis cs:1.2, 1 ) [penColor2,anchor=west] {$B$};
        \end{axis}
      \end{tikzpicture}\\
      \begin{tikzpicture}
        \begin{axis}[
          domain=-2:2,
          xmin=-2, xmax=2,
          ymin=-2, ymax=2,
          width=2.5in,
          axis lines =middle, xlabel=$x$, ylabel=$y$,
          every axis y label/.style={at=(current axis.above origin),anchor=south},
          every axis x label/.style={at=(current axis.right of origin),anchor=west},
          ]
	  \addplot [very thick, penColor3, smooth] {5*x^6-5*x^5-5*x^4+5*x^3+x^2 -.5};
          \node at (axis cs:1.2, 1 ) [penColor3,anchor=west] {$C$};
        \end{axis}
      \end{tikzpicture}
      &
      \begin{tikzpicture}
        \begin{axis}[
          domain=-2:2,
          xmin=-2, xmax=2,
          ymin=-2, ymax=2,
          width=2.5in,
          axis lines =middle, xlabel=$x$, ylabel=$y$,
          every axis y label/.style={at=(current axis.above origin),anchor=south},
          every axis x label/.style={at=(current axis.right of origin),anchor=west},
          ]
	  \addplot [very thick, penColor4, smooth,samples=100] {-5*x^6+5*x^5+5*x^4-5*x^3-x^2+1.5*x+1};
          \node at (axis cs:1.2, 1 ) [penColor4,anchor=west] {$D$};
        \end{axis}
      \end{tikzpicture}
    \end{tabular}
  \end{image}
  For each of the curves, determine if the polynomial has
  \textbf{even} or \textbf{odd} degree, and if the leading coefficient
  (the one next to the highest power of $x$) of the polynomial is
  \textbf{positive} or \textbf{negative}.
  \begin{explanation}\hfil
    \begin{itemize}
    \item Curve $A$ is defined by an
      \wordChoice{\choice{even}\choice[correct]{odd}} degree
      polynomial with a \wordChoice{\choice[correct]{positive}\choice{negative}}
      leading term.
    \item Curve $B$ is defined by an
      \wordChoice{\choice{even}\choice[correct]{odd}} degree
      polynomial with a
      \wordChoice{\choice{positive}\choice[correct]{negative}} leading
      term.
    \item Curve $C$ is defined by an
      \wordChoice{\choice[correct]{even}\choice{odd}} degree
      polynomial with a \wordChoice{\choice[correct]{positive}\choice{negative}}
      leading term.
    \item Curve $D$ is defined by an
      \wordChoice{\choice[correct]{even}\choice{odd}} degree
      polynomial with a \wordChoice{\choice{positive}\choice[correct]{negative}}
      leading term.
    \end{itemize}
  \end{explanation}
\end{example}


%% \section{Connections to inverse functions}

%% Polynomials only have inverse functions if they are
%% one-to-one. Pratically this means that polynomials of \textit{even}
%% degree are \textit{never} invertible since they are never
%% one-to-one. Odd degree polynomials may or may not be invertible. This
%% is worth repeating in the language of algebra, and we will write it as
%% a warning.

%% \begin{warning}
%%   Consider $f(x) = x^2$. Suppose we tried to find the inverse of
%%   $x^2$. In this case we plug $f^{-1}(x)$ into $f$ and write
%%   \begin{align*}
%%     f(f^{-1}(x)) &= \left(f^{-1}(x)\right)^2\\
%%     x &= \left(f^{-1}(x)\right)^2
%%   \end{align*}
%% at this point you may be tempted to say ``$f^{-1}(x) = \sqrt{x}$'' but
%% this is not correct. The correct next line is:
%% \[
%% f^{-1}(x) = \pm \sqrt{x},
%% \]
%% but this \textbf{is not a function} from the real numbers to the real
%% numbers.
%% \end{warning}


%% $x^\text{even}$ vs $x^\text{odd}$ and simple nth roots



\end{document}
